Live Functional Programming with Typed Holes

### 2018-10-23 calculus.tex
close #6

\parahead{Overview} The syntax of the core calculus given in Fig.~\ref{fig:hazelnut-live-syntax} consists of types and expressions with holes.
We distinguish between {external} expressions, $e$, and {internal} expressions, $d$.
External expressions correspond to programs as entered by the programmer
(see Sec.~\ref{sec:intro} for discussion of implicit, manual, semi-automated and fully automated hole entry methods).
Each well-typed external expression (see Sec.~\ref{sec:external-statics} below) expands to a well-typed internal expression (see Sec.~\ref{sec:expansion}) before it is evaluated (see Sec.~\ref{sec:evaluation}).
We take this approach, notably also taken in the ``redefinition'' of Standard ML by \citet{Harper00atype-theoretic}, because (1) the external language supports type inference and explicit type ascriptions, $\hexp : \htau$, but it is formally simpler to eliminate ascriptions and specify a type assignment system when defining the dynamic semantics; and
(2) we need additional syntactic machinery during evaluation for tracking hole closures and dynamic type casts.
This machinery is inserted by the expansion step, rather than entered explicitly by the programmer.
In this regard, the internal language is analogous to the cast calculus in the gradually typed lambda calculus \cite{DBLP:conf/snapl/SiekVCB15,Siek06a}, though as we will see the \HazelnutLive internal language goes beyond the cast calculus in several respects. We have mechanized these formal developments using the Agda proof assistant \cite{norell:thesis,norell2009dependently}
(see Sec.~\ref{sec:agda-mechanization}). Rule names in this section, e.g. \rulename{SVar}, correspond to variable names used in the mechanization. The \Hazel implementation substantially follows the formal specification of \Hazelnut (for the editor component) and \HazelnutLive (for the semantics), and we can formally state a continuity invariant for a putative combined calculus (see Sec.~\ref{sec:implementation}).

@@

\vspace{-4px}
\subsection{Expansion}
\label{sec:expansion}
\vspace{-1px}

\input{fig-expansion}
\input{fig-typing-dexp}

Each well-typed external expression~$e$ expands to a well-typed internal expression~$d$, for evaluation.
%
\Figref{fig:expansion} specifies expansion, and \Figref{fig:hasType} specifies type assignment for internal expressions.
%
% \Secref{sec:evaluation} discusses internal expression evaluation.

As with the type system for the external language (above),
we specify expansion bidirectionally \cite{DBLP:conf/ppdp/FerreiraP14}.
%
The synthetic expansion judgement~$\expandSyn{\hGamma}{\hexp}{\htau}{\dexp}{\Delta}$
%
produces an expansion~$d$ and a hole context~$\hDelta$ when synthesizing type $\htau$ for $\hexp$.
%
%We say more about hole contexts, which are used in the type assignment judgement, $\hasType{\Delta}{\hGamma}{d}{\htau}$, below.
We describe hole contexts, which serve as ``inputs'' to the type assignment judgement~$\hasType{\Delta}{\hGamma}{d}{\htau}$, further below.
%
The analytic expansion judgement~$\expandAna{\hGamma}{\hexp}{\htau}{\dexp}{\htau'}{\Delta}$, produces an expansion~$d$ of type~$\htau'$, and a hole context~$\hDelta$, when checking~$\hexp$ against~$\htau$.
%
The following theorem establishes that expansions are well-typed and in the analytic case that the assigned type, $\htau'$, is consistent with provided type, $\htau$.
%
\begin{thm}[Typed Expansion]\label{thm:typed-expansion} ~
  \begin{enumerate}[nolistsep]
    \item
      If $\expandSyn{\hGamma}{\hexp}{\htau}{\dexp}{\Delta}$
      then $\hasType{\Delta}{\hGamma}{\dexp}{\htau}$.
    \item
      If $\expandAna{\hGamma}{\hexp}{\htau}{\dexp}{\htau'}{\Delta}$
      then $\tconsistent{\htau}{\htau'}$ and $\hasType{\Delta}{\hGamma}{\dexp}{\htau'}$.
  \end{enumerate}
\end{thm}
\noindent
%
%The reason analytic expansion produces an expansion of consistent
%type is because the subsumption rule, as previously discussed, allows
%us to check an external expression against any type consistent with
%the type the expression actually synthesizes, whereas every internal
%expression can be assigned at most one type, i.e. the following
%standard unicity property holds of the type assignment system.
%
The reason that $\htau'$ is only consistent with the provided type $\htau$ is because
%
 the subsumption rule permits us
to check an external expression against any type consistent with the
type that the expression \emph{actually} synthesizes, whereas every internal
expression can be assigned at most one type, i.e. the following
standard unicity property holds of the type assignment system.
%
\begin{thm}[Type Assignment Unicity]
  If $\hasType{\Delta}{\hGamma}{\dexp}{\htau}$
  and $\hasType{\Delta}{\hGamma}{\dexp}{\htau'}$
  then $\htau=\htau'$.
\end{thm}
\noindent
Consequently, analytic expansion reports the type actually assigned to the expansion it produces.
%
For example, we can derive that $\expandAna{\hGamma}{c}{\tehole}{c}{b}{\emptyset}$.% where $\emptyset$ is the empty hole context.

Before describing the rules in detail, let us state two other guiding theorems.
The following theorem establishes that an expansion exists for every well-typed external expression.
The mechanization also establishes that when an expansion exists, it is unique (not shown).
 \begin{thm}[Expandability] \label{thm:expandability}~
  \begin{enumerate}[nolistsep]
    \item
      If $\hsyn{\hGamma}{\hexp}{\htau}$
      then $\expandSyn{\hGamma}{\hexp}{\htau}{\dexp}{\Delta}$
      for some $\dexp$ and $\Delta$.
    \item
      If $\hana{\hGamma}{\hexp}{\htau}$
      then $\expandAna{\hGamma}{\hexp}{\htau}{\dexp}{\htau'}{\Delta}$
      for some $\dexp$ and $\htau'$ and $\Delta$.
  \end{enumerate}
\end{thm}
% \noindent
% The following theorem establishes that when an expansion exists, it is unique.
% \begin{thm}[Expansion Unicity] \label{thm:expansion-unicity}~
%   % \begin{enumerate}[nolistsep]
%   %   \item
%       If $\expandSyn{\hGamma}{\hexp}{\htau}{\dexp}{\Delta}$
%       and $\expandSyn{\hGamma}{\hexp}{\htau'}{\dexp'}{\Delta'}$
%       then $\htau=\htau'$ and $\dexp=\dexp'$ and $\Delta=\Delta'$.
%   %   \item
%   %     If $\expandAna{\hGamma}{\hexp}{\htau_1}{\dexp}{\htau_2}{\Delta}$
%   %     and $\expandAna{\hGamma}{\hexp}{\htau_1}{\dexp'}{\htau_2'}{\Delta'}$
%   %     then $\dexp=\dexp'$ and $\htau_2=\htau_2'$ and $\Delta=\Delta'$.
%   % \end{enumerate}
% \end{thm}
% \noindent
The following theorem establishes that expansion generalizes external typing.
\begin{thm}[Expansion Generality] \label{thm:expansion-generality}~
  \begin{enumerate}[nolistsep]
    \item
      If $\expandSyn{\hGamma}{\hexp}{\htau}{\dexp}{\Delta}$
      then $\hsyn{\hGamma}{\hexp}{\htau}$.
    \item
      If $\expandAna{\hGamma}{\hexp}{\htau}{\dexp}{\htau'}{\Delta}$
      then $\hana{\hGamma}{\hexp}{\htau}$.
  \end{enumerate}
\end{thm}

The rules governing expansion of constants, variables and lambda expressions
--- Rules \rulename{ESConst}, \rulename{ESVar}, \rulename{ESLam} and \rulename{EALam}
--- mirror the corresponding type assignment rules
--- Rules \rulename{TAConst}, \rulename{TAVar} and \rulename{TALam}
--- and in turn, the corresponding bidirectional typing rules from Fig.~\ref{fig:bidirectional-typing}.
%
% Consequently, the corresponding cases of
% Theorem~\ref{thm:typed-expansion}, Theorem~\ref{thm:expandability} and
% Theorem~\ref{thm:expansion-generality} are straightforward.
%
To support type assignment, all lambdas in the internal language are
half-annotated---Rule \rulename{EALam} inserts the annotation when expanding an
unannotated external lambda based on the given type.
%
The rules governing hole expansion, and the rules that perform \emph{cast insertion}---those governing
function application and type ascription---are more interesting. Let us consider each of these two groups of rules in turn in Sec.~\ref{sec:hole-expansion} and Sec.~\ref{sec:cast-insertion}, respectively.

\subsubsection{Hole Expansion}\label{sec:hole-expansion}
Rules \rulename{ESEHole}, \rulename{ESNEHole}, \rulename{EAEHole} and \rulename{EANEHole} govern the expansion of empty and non-empty expression holes to empty and non-empty \emph{hole closures}, $\dehole{u}{\sigma}{}$ and $\dhole{\dexp}{u}{\sigma}{}$.
%
The hole name~$u$ on a hole closure identifies the external hole to which the hole closure corresponds.
%
While we assume each hole name to be unique in the external language, once evaluation begins, there may be multiple hole closures with the same name due to substitution.
%
For example, the result from Fig.~\ref{fig:grades-example} shows three closures for the hole named 1.
%
There, we numbered each hole closure for a given hole sequentially, \li{1:1}, \li{1:2} and \li{1:3}, but this is strictly for the sake of presentation, so we omit hole closure numbers from the core calculus.

%
For each hole, $u$, in an external expression, the hole context generated by expansion, $\Delta$, contains a hypothesis of the form~$\Dbinding{u}{\hGamma}{\htau}$, which records the hole's type, $\tau$, and the typing context, $\Gamma$, from where it appears in the original expression.\footnote{
We use a hole context, rather than recording the typing context and type directly on each hole closure, to ensure that all closures for a hole name have the same typing context and type.}
%
We borrow this hole context notation from contextual modal type theory (CMTT) \cite{Nanevski2008}, identifying hole names with metavariables and hole contexts with modal contexts (we say more about the connection with CMTT below).
%
% Each hole expansion rule records the ``current'' typing context under which the hole is expanded.
%
In the synthetic hole expansion rules~\rulename{ESEHole} and~\rulename{ESNEHole}, the generated hole context assigns the hole type~$\tehole$ to hole name~$u$, as in the external typing rules.
%
However, the first two premises of the expansion subsumption rule~\rulename{EASubsume} disallow the use of subsumption for holes in analytic position.
%
Instead, we employ separate analytic rules~\rulename{EAEHole} and~\rulename{EANEHole}, which each record the checked type~$\tau$ in the hole context.
%
Consequently, we can use type assignment for the internal language --- the type assignment rules \rulename{TAEHole} and \rulename{TANEHole} in Fig.~\ref{fig:hasType} assign a hole closure for hole name~$u$ the corresponding type from the hole context.

Each hole closure also has an associated environment~$\sigma$ which consists of a finite substitution of the form $[d_1/x_1, ~\cdots, d_n/x_n]$ for $n \geq 0$.
%
The closure environment keeps a record of the substitutions that occur around the hole as evaluation occurs.
%
Initially, when no evaluation has yet occurred, the hole expansion rules generate the identity substitution for the typing context associated with hole name~$u$ in hole context~$\Delta$, which we notate $\idof{\hGamma}$, and define as follows.
%
\begin{defn}[Identity Substitution] $\idof{x_1 : \tau_1, ~\cdots, x_n : \tau_n} = [x_1/x_1, ~\cdots, x_n/x_n]$
\end{defn}
\noindent
The type assignment rules for hole closures,~\rulename{TAEHole} and \rulename{TANEHole}, each require that the hole closure environment~$\sigma$ be consistent with the corresponding typing context, written as $\hasType{\Delta}{\hGamma}{\sigma}{\hGamma'}$.
%
Formally, we define this relation in terms of type assignment as follows:
\begin{defn}[Substitution Typing]
$\hasType{\Delta}{\hGamma}{\sigma}{\hGamma'}$ iff $\domof{\sigma} = \domof{\hGamma'}$ and for each $x : \htau \in \hGamma'$ we have that $d/x \in \sigma$ and $\hasType{\Delta}{\hGamma}{d}{\tau}$.
\end{defn}
\noindent
It is easy to verify that the identity substitution satisfies this requirement, i.e. that $\hasType{\Delta}{\hGamma}{\idof{\hGamma}}{\hGamma}$.

@@

Expansion inserts casts at function applications and ascriptions.
%
The latter is more straightforward: Rule~\rulename{ESAsc}
in \Figref{fig:expandSyn} inserts a cast from the assigned type to the
ascribed type.
%
Theorem~\ref{thm:typed-expansion} inductively ensures that the two
types are consistent.
%
We include ascription for exposition purposes---this form is derivable
by using application together with the half-annotated identity, $e
: \tau = \hap{(\halam{x}{\htau}{x})}{e}$; as such, application
expansion, discussed below, is more general.

Rule~\rulename{ESAp} expands function applications.
%
To understand the rule, consider the expansion of external
expression~$\hap{(\halam{x}{\tehole}{\hap{x}{c}})}{c}$, the example
discussed above:
\[
        \hap{\dcasttwo{
        (\halam{x}{\tehole}{\underbrace{
                \hap{\dcasttwo{x}{\tehole}{\tarr{\tehole}{\tehole}}}
                {\dcasttwo{c}{b}{\tehole}}
                }_{\textrm{expansion of function body}~x(c)}
        }
        )}{\tarr{\tehole}{\tehole}}{
           \tarr{\tehole}{\tehole}}
           }
           {\dcasttwo{c}{b}{\tehole}}
\]
Consider the (indicated) function body,
%
where expansion inserts a cast on both the function expression~$x$ and its argument~$c$.
%
Together, these casts for~$x$ and~$c$ permit assigning a type to the
function body according to the rules in \Figref{fig:hasType}, where we
could not do so under the same context without casts.
%
We separately consider the expansions of~$x$ and of~$c$.

First, consider the function position of this application, here variable~$x$.
%
Without any cast, the type of variable~$x$ is the hole type~$\tehole$;
however, the inserted cast on~$x$ permits treating it as though it has
arrow type $\tarr{\tehole}{\tehole}$.
%
The first three premises of Rule~\rulename{ESAp} accomplish this
%
by first synthesizing a type for the function expression, here
$\tehole$, then
%
by determining the matched arrow type~$\tarr{\tehole}{\tehole}$, and
finally,
%
by performing analytic expansion on the function expression with this
matched arrow type.
%
The resulting expansion has some type~$\tau_1'$ consistent with the matched arrow type.
%
In this case, because the subexpression~$x$ is a variable, analytic
expansion goes through subsumption so that type~$\tau_1'$ is
simply~$\tehole$.
%
The conclusion of the rule inserts the corresponding cast.
%
We go through type synthesis, \emph{then} analytic expansion, so that the hole
context records the matched arrow type for holes in function position,
rather than the type~$\tehole$ for all such holes, as would be the case
in a variant of this rule using synthetic expansion for the function
expression.

Next, consider the application's argument, here constant~$c$.
%
The conclusion of Rule~\rulename{ESAp} inserts the cast on the argument's
expansion, from the type it is assigned by the final premise of the
rule~(type~$b$ here), to the argument type of the matched arrow type
of the function expression~(type~$\tehole$ here).

The example's second, outermost application goes through the same
application expansion rule.
%
In this case, the cast on the function is the identity cast for
$\tarr{\tehole}{\tehole}$.
%
For simplicity, we do not attempt to avoid the insertion of identity
casts in the core calculus; these will simply never fail during
evaluation.
%
However, it is safe in practice to eliminate such identity casts during expansion,
and some formal accounts of gradual typing do so by defining three
application expansion rules, including the original account of \citet{Siek06a}.

@@

To recap, the result of expansion is a well-typed internal expression with appropriately initialized hole closures and casts.
This section specifies the dynamic semantics of \HazelnutLive as a ``small-step'' transition system over internal expressions equipped with a meaningful notion of type safety even for incomplete programs, i.e. expressions typed under a non-empty hole context, $\Delta$.
We establish that evaluation does not stop immediately when it encounters a hole, nor when a cast fails, by precisely characterizing when evaluation \emph{does} stop.
We also establish that an essentially standard notion of type safety holds when running complete programs.

@@

\subsubsection{Complete Programs}
%
Although this paper focuses on running \emph{incomplete} programs, it helps
to know that the necessary machinery does not interfere
with running \emph{complete} programs, i.e. those
with no type or expression holes.
%
% Appendix~\ref{sec:additional-defns}
The \appendixName defines the predicates~$\isComplete{e}$, $\isComplete{\htau}$, $\isComplete{\hGamma}$ and~$\isComplete{d}$.
%
Of note, failed casts cannot appear in complete internal expressions.
%
The following theorem establishes that expansion preserves program completeness.

\begin{thm}[Complete Expansion] 
% ~
  % \begin{enumerate}[nolistsep]
      If $\isComplete{\hGamma}$ and $\isComplete{\hexp}$
      and $\expandSyn{\hGamma}{\hexp}{\htau}{\dexp}{\Delta}$
      then $\isComplete{\htau}$ and $\isComplete{\dexp}$ and $\Delta = \emptyset$.
%     \item
%       If $\isComplete{\hGamma}$ and $\isComplete{\hexp}$
% and $\isComplete{\htau}$
%       and $\expandAna{\hGamma}{\hexp}{\htau}{\dexp}{\htau'}{\Delta}$
%       then $\isComplete{\dexp}$ and $\isComplete{\htau'}$ and $\Delta=\emptyset$
  % \end{enumerate}
\end{thm}

>>>>

\parahead{Overview} The syntax of the core calculus given in
Fig.~\ref{fig:hazelnut-live-syntax} consists of types and expressions with
holes.  We distinguish between {external} expressions, $e$, and {internal}
expressions, $d$.  External expressions correspond to programs as entered
by the programmer (see Sec.~\ref{sec:intro} for discussion of implicit,
manual, semi-automated and fully automated hole entry methods).  Each
well-typed external expression (see Sec.~\ref{sec:external-statics} below)
elaborates to a well-typed internal expression (see
Sec.~\ref{sec:elaboration}) before it is evaluated (see
Sec.~\ref{sec:evaluation}).  We take this approach, notably also taken in
the ``redefinition'' of Standard ML by \citet{Harper00atype-theoretic},
because (1) the external language supports type inference and explicit type
ascriptions, $\hexp : \htau$, but it is formally simpler to eliminate
ascriptions and specify a type assignment system when defining the dynamic
semantics; and (2) we need additional syntactic machinery during evaluation
for tracking hole closures and dynamic type casts.  This machinery is
inserted by elaboration, rather than entered explicitly by the programmer.
In this regard, the internal language is analogous to the cast calculus in
the gradually typed lambda calculus
\cite{DBLP:conf/snapl/SiekVCB15,Siek06a}, though as we will see the
\HazelnutLive internal language goes beyond the cast calculus in several
respects. We have mechanized these formal developments using the Agda proof
assistant \cite{norell:thesis,norell2009dependently} (see
Sec.~\ref{sec:agda-mechanization}). Rule names in this section,
e.g. \rulename{SVar}, correspond to variable names used in the
mechanization. The \Hazel implementation substantially follows the formal
specification of \Hazelnut (for the editor component) and \HazelnutLive
(for the semantics), and we can formally state a continuity invariant for a
putative combined calculus (see Sec.~\ref{sec:implementation}).

@@

\vspace{-4px}
\subsection{Elaboration}
\label{sec:elaboration}
\vspace{-1px}

\input{fig-elab}
\input{fig-typing-dexp}

Each well-typed external expression~$e$ elaborates to a well-typed internal
expression~$d$, for evaluation.
%
\Figref{fig:elaboration} specifies elaboration, and \Figref{fig:hasType}
specifies type assignment for internal expressions.
%
% \Secref{sec:evaluation} discusses internal expression evaluation.

As with the type system for the external language (above), we specify
elaboration bidirectionally \cite{DBLP:conf/ppdp/FerreiraP14}.
%
The synthetic elaboration
judgement~$\elabSyn{\hGamma}{\hexp}{\htau}{\dexp}{\Delta}$
%
produces an elaboration~$d$ and a hole context~$\hDelta$ when synthesizing
type $\htau$ for $\hexp$.
%
%We say more about hole contexts, which are used in the type assignment judgement, $\hasType{\Delta}{\hGamma}{d}{\htau}$, below.
We describe hole contexts, which serve as ``inputs'' to the type assignment
judgement~$\hasType{\Delta}{\hGamma}{d}{\htau}$, further below.
%
The analytic elaboration
judgement~$\elabAna{\hGamma}{\hexp}{\htau}{\dexp}{\htau'}{\Delta}$,
produces an elaboration~$d$ of type~$\htau'$, and a hole context~$\hDelta$,
when checking~$\hexp$ against~$\htau$.
%
The following theorem establishes that elaborations are well-typed and in
the analytic case that the assigned type, $\htau'$, is consistent with
provided type, $\htau$.
%
\begin{thm}[Typed Elaboration]\label{thm:typed-elaboration} ~
  \begin{enumerate}[nolistsep]
    \item
      If $\elabSyn{\hGamma}{\hexp}{\htau}{\dexp}{\Delta}$
      then $\hasType{\Delta}{\hGamma}{\dexp}{\htau}$.
    \item
      If $\elabAna{\hGamma}{\hexp}{\htau}{\dexp}{\htau'}{\Delta}$ then
      $\tconsistent{\htau}{\htau'}$ and
      $\hasType{\Delta}{\hGamma}{\dexp}{\htau'}$.
  \end{enumerate}
\end{thm}
\noindent
%
%The reason analytic expansion produces an expansion of consistent
%type is because the subsumption rule, as previously discussed, allows
%us to check an external expression against any type consistent with
%the type the expression actually synthesizes, whereas every internal
%expression can be assigned at most one type, i.e. the following
%standard unicity property holds of the type assignment system.
%
The reason that $\htau'$ is only consistent with the provided type $\htau$ is because
%
the subsumption rule permits us to check an external expression against any
type consistent with the type that the expression \emph{actually}
synthesizes, whereas every internal expression can be assigned at most one
type, i.e. the following standard unicity property holds of the type
assignment system.
%
\begin{thm}[Type Assignment Unicity]
  If $\hasType{\Delta}{\hGamma}{\dexp}{\htau}$
  and $\hasType{\Delta}{\hGamma}{\dexp}{\htau'}$
  then $\htau=\htau'$.
\end{thm}
\noindent
Consequently, analytic elaboration reports the type actually assigned to
the elaboration it produces.
%
For example, we can derive that
$\elabAna{\hGamma}{c}{\tehole}{c}{b}{\emptyset}$.
% where $\emptyset$ is the empty hole context.

Before describing the rules in detail, let us state two other guiding
theorems.  The following theorem establishes that every well-typed external
expression can be elaborated.
 \begin{thm}[Elaborability] \label{thm:elaborability}~
  \begin{enumerate}[nolistsep]
    \item
      If $\hsyn{\hGamma}{\hexp}{\htau}$
      then $\elabSyn{\hGamma}{\hexp}{\htau}{\dexp}{\Delta}$
      for some $\dexp$ and $\Delta$.
    \item
      If $\hana{\hGamma}{\hexp}{\htau}$
      then $\elabAna{\hGamma}{\hexp}{\htau}{\dexp}{\htau'}{\Delta}$
      for some $\dexp$ and $\htau'$ and $\Delta$.
  \end{enumerate}
 \end{thm}

% \noindent
% The following theorem establishes that when an expansion exists, it is unique.
% \begin{thm}[Expansion Unicity] \label{thm:expansion-unicity}~
%   % \begin{enumerate}[nolistsep]
%   %   \item
%       If $\elabSyn{\hGamma}{\hexp}{\htau}{\dexp}{\Delta}$
%       and $\elabSyn{\hGamma}{\hexp}{\htau'}{\dexp'}{\Delta'}$
%       then $\htau=\htau'$ and $\dexp=\dexp'$ and $\Delta=\Delta'$.
%   %   \item
%   %     If $\elabAna{\hGamma}{\hexp}{\htau_1}{\dexp}{\htau_2}{\Delta}$
%   %     and $\elabAna{\hGamma}{\hexp}{\htau_1}{\dexp'}{\htau_2'}{\Delta'}$
%   %     then $\dexp=\dexp'$ and $\htau_2=\htau_2'$ and $\Delta=\Delta'$.
%   % \end{enumerate}
% \end{thm}
\noindent
The following theorem establishes that elaboration generalizes external
typing.
\begin{thm}[Elaboration Generality] \label{thm:elaboration-generality}~
  \begin{enumerate}[nolistsep]
    \item
      If $\elabSyn{\hGamma}{\hexp}{\htau}{\dexp}{\Delta}$
      then $\hsyn{\hGamma}{\hexp}{\htau}$.
    \item
      If $\elabAna{\hGamma}{\hexp}{\htau}{\dexp}{\htau'}{\Delta}$
      then $\hana{\hGamma}{\hexp}{\htau}$.
  \end{enumerate}
\end{thm}

We also establish that the elaboration produces unique results.
\begin{thm}[Elaboration Unicity] \label{thm:expansion-unicity}~
  \begin{enumerate}[nolistsep]
  \item If $\elabSyn{\hGamma}{\hexp}{\htau_1}{\dexp{}_1}{\Delta_1}$ and
    $\elabSyn{\hGamma}{\hexp}{\htau_2}{\dexp{}_2}{\Delta_2}$, then
    $\htau_1 = \htau_2$ and $\dexp{}_1 = \dexp{}_2$ and $\Delta_1 =
    \Delta_2$.
  \item If
    $\elabAna{\hGamma}{\hexp}{\htau}{\dexp{}_1}{\htau_1}{\Delta_1}$ and
    $\elabAna{\hGamma}{\hexp}{\htau}{\dexp{}_2}{\htau_2}{\Delta_2}$, then
    $\htau_1 = \htau_2$ and $\dexp{}_1 = \dexp{}_2$ and $\Delta_1 =
    \Delta_2$
  \end{enumerate}
\end{thm}

The rules governing elaboration of constants, variables and lambda
expressions---Rules \rulename{ESConst}, \rulename{ESVar}, \rulename{ESLam}
and \rulename{EALam}---mirror the corresponding type assignment rules---
Rules \rulename{TAConst}, \rulename{TAVar} and \rulename{TALam}---and in
turn, the corresponding bidirectional typing rules from
Fig.~\ref{fig:bidirectional-typing}.
%
% Consequently, the corresponding cases of
% Theorem~\ref{thm:typed-expansion}, Theorem~\ref{thm:expandability} and
% Theorem~\ref{thm:expansion-generality} are straightforward.
%
To support type assignment, all lambdas in the internal language are
half-annotated---Rule \rulename{EALam} inserts the annotation when
elaborating an unannotated external lambda based on the given type.
%
The rules governing hole elaboration, and the rules that perform \emph{cast
  insertion}---those governing function application and type
ascription---are more interesting. Let us consider each of these two groups
of rules in turn in Sec.~\ref{sec:hole-elaboration} and
Sec.~\ref{sec:cast-insertion}, respectively.

\subsubsection{Hole Elaboration}\label{sec:hole-elaboration}
Rules \rulename{ESEHole}, \rulename{ESNEHole}, \rulename{EAEHole} and
\rulename{EANEHole} govern the elaboration of empty and non-empty
expression holes to empty and non-empty \emph{hole closures},
$\dehole{u}{\sigma}{}$ and $\dhole{\dexp}{u}{\sigma}{}$.
%
The hole name~$u$ on a hole closure identifies the external hole to which
the hole closure corresponds.
%
While we assume each hole name to be unique in the external language, once
evaluation begins, there may be multiple hole closures with the same name
due to substitution.
%
For example, the result from Fig.~\ref{fig:grades-example} shows three
closures for the hole named 1.
%
There, we numbered each hole closure for a given hole sequentially,
\li{1:1}, \li{1:2} and \li{1:3}, but this is strictly for the sake of
presentation, so we omit hole closure numbers from the core calculus.

%
For each hole, $u$, in an external expression, the hole context generated
by elaboration, $\Delta$, contains a hypothesis of the
form~$\Dbinding{u}{\hGamma}{\htau}$, which records the hole's type, $\tau$,
and the typing context, $\Gamma$, from where it appears in the original
expression.\footnote{ We use a hole context, rather than recording the
  typing context and type directly on each hole closure, to ensure that all
  closures for a hole name have the same typing context and type.}
%
We borrow this hole context notation from contextual modal type theory
(CMTT) \cite{Nanevski2008}, identifying hole names with metavariables and
hole contexts with modal contexts (we say more about the connection with
CMTT below).
%
% Each hole expansion rule records the ``current'' typing context under which the hole is expanded.
%
In the synthetic hole elaboration rules~\rulename{ESEHole}
and~\rulename{ESNEHole}, the generated hole context assigns the hole
type~$\tehole$ to hole name~$u$, as in the external typing rules.
%
However, the first two premises of the elaboration subsumption
rule~\rulename{EASubsume} disallow the use of subsumption for holes in
analytic position.
%
Instead, we employ separate analytic rules~\rulename{EAEHole}
and~\rulename{EANEHole}, which each record the checked type~$\tau$ in the
hole context.
%
Consequently, we can use type assignment for the internal language --- the
type assignment rules \rulename{TAEHole} and \rulename{TANEHole} in
Fig.~\ref{fig:hasType} assign a hole closure for hole name~$u$ the
corresponding type from the hole context.

Each hole closure also has an associated environment~$\sigma$ which
consists of a finite substitution of the form $[d_1/x_1, ~\cdots, d_n/x_n]$
for $n \geq 0$.
%
The closure environment keeps a record of the substitutions that occur around the hole as evaluation occurs.
%
Initially, when no evaluation has yet occurred, the hole elaboration rules
generate the identity substitution for the typing context associated with
hole name~$u$ in hole context~$\Delta$, which we notate $\idof{\hGamma}$,
and define as follows.
%
\begin{defn}[Identity Substitution] $\idof{x_1 : \tau_1, ~\cdots, x_n : \tau_n} = [x_1/x_1, ~\cdots, x_n/x_n]$
\end{defn}
\noindent
The type assignment rules for hole closures,~\rulename{TAEHole} and \rulename{TANEHole}, each require that the hole closure environment~$\sigma$ be consistent with the corresponding typing context, written as $\hasType{\Delta}{\hGamma}{\sigma}{\hGamma'}$.
%
Formally, we define this relation in terms of type assignment as follows:
\begin{defn}[Substitution Typing]
$\hasType{\Delta}{\hGamma}{\sigma}{\hGamma'}$ iff $\domof{\sigma} = \domof{\hGamma'}$ and for each $x : \htau \in \hGamma'$ we have that $d/x \in \sigma$ and $\hasType{\Delta}{\hGamma}{d}{\tau}$.
\end{defn}
\noindent
It is easy to verify that the identity substitution satisfies this requirement, i.e. that $\hasType{\Delta}{\hGamma}{\idof{\hGamma}}{\hGamma}$.

@@

Elaboration inserts casts at function applications and ascriptions.
%
The latter is more straightforward: Rule~\rulename{ESAsc}
in \Figref{fig:elabSyn} inserts a cast from the assigned type to the
ascribed type.
%
Theorem~\ref{thm:typed-elaboration} inductively ensures that the two types
are consistent.
%
We include ascription for exposition purposes---this form is derivable
by using application together with the half-annotated identity, $e
: \tau = \hap{(\halam{x}{\htau}{x})}{e}$; as such, application
elaboration, discussed below, is more general.

Rule~\rulename{ESAp} elaborates function applications.
%
To understand the rule, consider the elaboration of external
expression~$\hap{(\halam{x}{\tehole}{\hap{x}{c}})}{c}$, the example
discussed above:
\[
        \hap{\dcasttwo{
        (\halam{x}{\tehole}{\underbrace{
                \hap{\dcasttwo{x}{\tehole}{\tarr{\tehole}{\tehole}}}
                {\dcasttwo{c}{b}{\tehole}}
                }_{\textrm{elaboration of function body}~x(c)}
        }
        )}{\tarr{\tehole}{\tehole}}{
           \tarr{\tehole}{\tehole}}
           }
           {\dcasttwo{c}{b}{\tehole}}
\]
Consider the (indicated) function body,
%
where elaboration inserts a cast on both the function expression~$x$ and
its argument~$c$.
%
Together, these casts for~$x$ and~$c$ permit assigning a type to the
function body according to the rules in \Figref{fig:hasType}, where we
could not do so under the same context without casts.
%
We separately consider the elaborations of~$x$ and of~$c$.

First, consider the function position of this application, here variable~$x$.
%
Without any cast, the type of variable~$x$ is the hole type~$\tehole$;
however, the inserted cast on~$x$ permits treating it as though it has
arrow type $\tarr{\tehole}{\tehole}$.
%
The first three premises of Rule~\rulename{ESAp} accomplish this
%
by first synthesizing a type for the function expression, here
$\tehole$, then
%
by determining the matched arrow type~$\tarr{\tehole}{\tehole}$, and
finally,
%
by performing analytic elaboration on the function expression with this
matched arrow type.
%
The resulting elaboration has some type~$\tau_1'$ consistent with the
matched arrow type.
%
In this case, because the subexpression~$x$ is a variable, analytic
elaboration goes through subsumption so that type~$\tau_1'$ is
simply~$\tehole$.
%
The conclusion of the rule inserts the corresponding cast.
%
We go through type synthesis, \emph{then} analytic elaboration, so that the
hole context records the matched arrow type for holes in function position,
rather than the type~$\tehole$ for all such holes, as would be the case in
a variant of this rule using synthetic elaboration for the function
expression.

Next, consider the application's argument, here constant~$c$.
%
The conclusion of Rule~\rulename{ESAp} inserts the cast on the argument's
elaboration, from the type it is assigned by the final premise of the
rule~(type~$b$ here), to the argument type of the matched arrow type of the
function expression~(type~$\tehole$ here).

The example's second, outermost application goes through the same
application elaboration rule.
%
In this case, the cast on the function is the identity cast for
$\tarr{\tehole}{\tehole}$.
%
For simplicity, we do not attempt to avoid the insertion of identity
casts in the core calculus; these will simply never fail during
evaluation.
%
However, it is safe in practice to eliminate such identity casts during
elaboration, and some formal accounts of gradual typing do so by defining
three application elaboration rules, including the original account of
\citet{Siek06a}.

@@

To recap, the result of elaboration is a well-typed internal expression
with appropriately initialized hole closures and casts.  This section
specifies the dynamic semantics of \HazelnutLive as a ``small-step''
transition system over internal expressions equipped with a meaningful
notion of type safety even for incomplete programs, i.e. expressions typed
under a non-empty hole context, $\Delta$.  We establish that evaluation
does not stop immediately when it encounters a hole, nor when a cast fails,
by precisely characterizing when evaluation \emph{does} stop.  We also
establish that an essentially standard notion of type safety holds when
running complete programs.

@@

\subsubsection{Complete Programs}
%
Although this paper focuses on running \emph{incomplete} programs, it helps
to know that the necessary machinery does not interfere
with running \emph{complete} programs, i.e. those
with no type or expression holes.
%
% Appendix~\ref{sec:additional-defns}
The \appendixName{} defines the predicates~$\isComplete{e}$,
$\isComplete{\htau}$, $\isComplete{\hGamma}$ and~$\isComplete{d}$.
%
Of note, failed casts cannot appear in complete internal expressions.
%
The following theorem establishes that elaboration preserves program
completeness.

\begin{thm}[Complete Elaboration]
% ~
  % \begin{enumerate}[nolistsep]
      If $\isComplete{\hGamma}$ and $\isComplete{\hexp}$
      and $\elabSyn{\hGamma}{\hexp}{\htau}{\dexp}{\Delta}$
      then $\isComplete{\htau}$ and $\isComplete{\dexp}$ and $\Delta = \emptyset$.
%     \item
%       If $\isComplete{\hGamma}$ and $\isComplete{\hexp}$
% and $\isComplete{\htau}$
%       and $\elabAna{\hGamma}{\hexp}{\htau}{\dexp}{\htau'}{\Delta}$
%       then $\isComplete{\dexp}$ and $\isComplete{\htau'}$ and $\Delta=\emptyset$
  % \end{enumerate}
\end{thm}
